Methods for Projecting Population Growth

### 2018-05-07 final/tex/Paper.tex
Revision

Population growth is one of the many models of great importance to the scientific community. These models are used to predict the challenges and changes needed in the near and longterm future as the world's population inevitably grows. It is such an important subject that the UN frequently discusses new and better ways of modelling the future population. This has led to a number of competing models, each with its own strengths and weaknesses. While no model can be perfect, some have come relatively close to matching existing data and predict reasonable growth in the future.

In order to understand just a how a few of these models operate, this paper will explore 3 separate models: standard logistic growth (LG), logistic growth augmented with migration data (LGM), and a modified Lotka-Volterra model (MLV). Each model has its own strengths and weaknesses and perform differently in comparison to actual data provided by the United Nations World Bank.

@@

\begin{align}
\label{eq:log-growth}
	P(t) = \frac{K N_0 e^{\gamma t}}{K + N_0 (e^{\gamma t} - 1)}
\end{align}
with $K$ being the theoretical maximum sustainable population, $N_0$ the initial population at time $t_0$, $\gamma$ being the yearly growth rate, and $t$ being the time since $t_0$.

@@

While $N_0$ is trivially obtained from the dataset, $\gamma$ and $K$ must be calculated based on the dataset. These values are optimized utilizing the least squares method. This was used on data collected from the World Bank, which provides information dating back to 1960. The model was simulated on the world, United States, and South African populations (see section 3 for reasoning).

@@

These numbers seem rather reasonable, with rather tame growth rates and logically sound maximum populations. The growth rate of South Africa is larger than that of the world, which makes sense considering it is a rapidly developing country. On the opposite end, the growth rate for the U.S. is lower than that of the world due to the fact that it is already heavily developed and has far less room for growth.

@@

The logistic growth model, despite its limitations, consistently produces results that make sense while also fitting the data excellently. These results suggest that the modified Lotka-Volterra model is fundamentally flawed when working with economies that are not consistently growing. It seems to assume that an economy should always be growing and fails to account for any variances in that respect. More than that, the model requires a large amount of data to function correctly. One of the reasons why the model for the United States performed poorly, despite having a consistently growing economy, may be due to the lack of historical data. The original paper optimized the model using data dating back to 1850 whereas the simulations here only use data back to 1950 due to the lack of consistent sources.

With more data, this model would likely produce better results for the United States and similar countries. However, the model thrives on a larger scale, where fewer factors have a significant effect in population growth.

>>>>

Population growth is one of the many models of great importance to the scientific community. These models are used to predict the challenges and changes needed in the near and longterm future as the world's population inevitably grows. It is such an important subject that the UN frequently discusses new and better ways of modeling the future population. This has led to a number of competing models, each with its own strengths and weaknesses. While no model can be perfect, some have come relatively close to matching existing data and predict reasonable growth in the future.

In order to understand just a how a few of these models operate, this paper will explore 2 separate models: standard logistic growth (LG) and a modified Lotka-Volterra model (MLV). Each model has its own strengths and weaknesses and perform differently in comparison to actual data provided by the United Nations World Bank.

@@

\begin{align}
\label{eq:log-growth}
	N(t) = \frac{K N_0 e^{\gamma t}}{K + N_0 (e^{\gamma t} - 1)}
\end{align}
with $K$ being the theoretical maximum sustainable population, $N_0$ the initial population at time $t_0$, $\gamma$ being the yearly growth rate, and $t$ being the time since $t_0$.

@@

While $N_0$ is trivially obtained from the dataset, $\gamma$ and $K$ must be calculated based on the dataset. To do so, the least squares method is utilized on the equation:

\begin{align}
	\frac{dN}{dt} = \gamma - \frac{\gamma N(t)}{K}
\end{align}

This was used on data collected from the World Bank, which provides information dating back to 1960. The model was simulated on the world, United States, and South African populations (see section 3 for reasoning).

@@

These numbers seem reasonable, with rather tame growth rates and logically sound maximum populations. The growth rate of South Africa is larger than that of the world, which makes sense considering it is a rapidly developing country. On the opposite end, the growth rate for the U.S. is lower than that of the world due to the fact that it is already heavily developed and has far less room for growth.

@@

The logistic growth model, despite its limitations, consistently produces results that make sense while also fitting the data excellently. These results suggest that the modified Lotka-Volterra model is flawed when working on smaller systems, where other factors have a much large impact on population and economic growth. It seems to assume that an economy should always be growing and fails to account for any variances in that respect. More than that, the model requires a large amount of data to function correctly, something that is unfortunately not readily available in most cases. With more data, this model would likely produce better results for the United States and similar countries. However, the model thrives on a larger scale, where fewer factors have a significant effect in population growth.
